Draft_Application_Stochastic_Volatility

### 2021-08-09 Real Dataset SV.Rmd
grafici realised volatility

Once more, most algorithms used (aside from SIS and Liu-West) seem to provide indistinguishable results and, generally speaking, to underestimate the volatility (as indicated by the realised volatility proxy). The SIS algorithm appears completely inadequate to describe the "true" volatility process: as we argued in section XX, the degeneracy of its weights on the particles prevents it from delivering a good enough approximation of the target distribution (\textit{check whether this is the correct intuition}). On top of that, it does not seem the case that the SIS algorithm, by falling short of the SV target distribution, ends up fitting the proxy better.\footnote{As we will now see, this seems instead to be the case for our preferred algorithms, namely those adopting adaptive resampling rules: a better approximation of the SV target filtering distribution yields slightly higher RMSEs and MAEs than those resulting from less accurate MC approximations.}\
On the other hand, the Liu and West filter offers the best performance among all filters, at least visually. This suggests that, since the Liu and West particle filter is built on a different underlying model than the other filters considered, this model may offer a better representation of the data analyzed. Once more, these considerations hold for all of the three indices.\

GRAFICI

@@

Finally, we can see from Table XX the RMSE and MAE obtained with a number of particles equal to 50000, for the SP500 and the EuroStoxx50 indices. We consider the Bootstrap particle filter and the Auxiliary particle filter, which we employed in section XX; in addition to these, we add the Liu and West filter, which did not feature in that section of this work.\
In the case of the SP500 index, although the number of particles increases, the RMSE and MAE increase (although by very small amounts) for both the particle filters based on our usual specification of the SV model. Instead, for the Liu and West filter we see some improvements, although the small extent suggests that also the LW filter was already in the convergence phase with N=10000.\
In the case of the EuroStoxx50 index, while the Bootstrap filter increases both RMSE and MAE when N is raised to 50000, the APF shows improvements in its performance, according to both criteria, which instead disagree for the Liu and West filter. COMMENTI DA AGGIORNARE

>>>>

Once more, most algorithms used (aside from SIS and Liu-West) seem to provide indistinguishable results. The SIS algorithm appears completely inadequate to describe the "true" volatility process: as we argued in section XX, the degeneracy of its weights on the particles prevents it from delivering a good enough approximation of the target distribution (\textit{check whether this is the correct intuition}). On top of that, it does not seem the case that the SIS algorithm, by falling short of the SV target distribution, ends up fitting the proxy better.\footnote{As we will now see, this seems instead to be the case for our preferred algorithms, namely those adopting adaptive resampling rules: a better approximation of the SV target filtering distribution yields slightly higher RMSEs and MAEs than those resulting from less accurate MC approximations.}\
On the other hand, the Liu and West filter offers the best performance among all filters, at least visually. This suggests that, since the Liu and West particle filter is built on a different underlying model than the other filters considered, this model may offer a better representation of the data analyzed. Once more, these considerations hold for all of the three indices.\

```{r echo=FALSE, fig.align='center', fig.cap="Filtered volatility and realised volatility", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}

library(openxlsx)
library(ggplot2)
library(ggpubr)

realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
y<-realdatasetSPX[,2]
volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
x1<-volatilitySPX[,3]
realisedx1<-x1
timeframe<-c(1:length(y))

sv_bpf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 1))[,1]
sv_gpfopt1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 2))[,1]
sv_apf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 3))[,1]
sv_lw1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 4))[,1]
sv_bapf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 5))[,1]
sv_sis1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 6))[,1]
dfsv1<-data.frame(timeframe,y,realisedx1,sv_bpf1)
dfsv2<-data.frame(timeframe,y,realisedx1,sv_gpfopt1)
dfsv3<-data.frame(timeframe,y,realisedx1,sv_apf1)
dfsv4<-data.frame(timeframe,y,realisedx1,sv_lw1)
dfsv5<-data.frame(timeframe,y,realisedx1,sv_bapf1)
dfsv6<-data.frame(timeframe,y,realisedx1,sv_sis1)

Filtplot1<-function(dataframe,title){
  mean=sv_bpf1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Proxy" = "black",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Proxy" = 3,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot2<-function(dataframe,title){
  mean=sv_gpfopt1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Proxy" = "black",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Proxy" = 3,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot3<-function(dataframe,title){
  mean=sv_apf1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Proxy" = "black",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Proxy" = 3,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot4<-function(dataframe,title){
  mean=sv_lw1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Proxy" = "black",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Proxy" = 3,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot5<-function(dataframe,title){
  mean=sv_bapf1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Proxy" = "black",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Proxy" = 3,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot6<-function(dataframe,title){
  mean=sv_sis1
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Proxy" = "black",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Proxy" = 3,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}

plotproxy1<-Filtplot1(dfsv1,"Bootstrap Particle Filter SP500")
plotproxy2<-Filtplot2(dfsv2,"Opt Ker Guided Particle Filter SP500")
plotproxy3<-Filtplot3(dfsv3,"Auxiliary Particle Filter SP500")
plotproxy4<-Filtplot4(dfsv4,"Liu and West Filter SP500")
plotproxy5<-Filtplot5(dfsv5,"Always Resampling SP500")
plotproxy6<-Filtplot6(dfsv6,"No Resampling SP500")

plotproxy1
plotproxy2
plotproxy3
plotproxy4
plotproxy5
plotproxy6

realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 2)
y<-realdatasetSPX[,2]
volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 2)
x2<-volatilitySPX[,3]
realisedx2<-x2
timeframe<-c(1:length(y))

sv_bpf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 1))[,1]
sv_gpfopt2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 2))[,1]
sv_apf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 3))[,1]
sv_lw2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 4))[,1]
sv_bapf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 5))[,1]
sv_sis2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 6))[,1]
dfsv7<-data.frame(timeframe,y,realisedx2,sv_bpf2)
dfsv8<-data.frame(timeframe,y,realisedx2,sv_gpfopt2)
dfsv9<-data.frame(timeframe,y,realisedx2,sv_apf2)
dfsv10<-data.frame(timeframe,y,realisedx2,sv_lw2)
dfsv11<-data.frame(timeframe,y,realisedx2,sv_bapf2)
dfsv12<-data.frame(timeframe,y,realisedx2,sv_sis2)

Filtplot7<-function(dataframe,title){
  mean=sv_bpf2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Proxy" = "black",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Proxy" = 3,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot8<-function(dataframe,title){
  mean=sv_gpfopt2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Proxy" = "black",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Proxy" = 3,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot9<-function(dataframe,title){
  mean=sv_apf2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Proxy" = "black",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Proxy" = 3,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot10<-function(dataframe,title){
  mean=sv_lw2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Proxy" = "black",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Proxy" = 3,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot11<-function(dataframe,title){
  mean=sv_bapf2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Proxy" = "black",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Proxy" = 3,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}
Filtplot12<-function(dataframe,title){
  mean=sv_sis2
  mean1=exp(mean/2)
  dataframe<-data.frame(dataframe,mean1)

  ggplot(dataframe,aes(x=timeframe))+
    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
    scale_color_manual("",
                       values=c("Proxy" = "black",
                                "Filtered Volatility"= "red"))+
    scale_linetype_manual("",
                          values=c("Proxy" = 3,
                                   "Filtered Volatility"=1))+
    labs(x="Time",
         y="")+
    ggtitle(title)+
    theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
          legend.background = element_rect(fill = "white", colour = "gray30")) +
    theme(plot.title = element_text(hjust = 0.5))
}

plotproxy7<-Filtplot7(dfsv7,"Bootstrap Particle Filter STOXX")
plotproxy8<-Filtplot8(dfsv8,"Opt Ker Guided Particle Filter STOXX")
plotproxy9<-Filtplot9(dfsv9,"Auxiliary Particle Filter STOXX")
plotproxy10<-Filtplot10(dfsv10,"Liu and West Filter STOXX")
plotproxy11<-Filtplot11(dfsv11,"Always Resampling STOXX")
plotproxy12<-Filtplot12(dfsv12,"No Resampling STOXX")

plotproxy7
plotproxy8
plotproxy9
plotproxy10
plotproxy11
plotproxy12

@@

Finally, we can see from Table XX the RMSE and MAE obtained with a number of particles equal to 50000, for the SP500 and the EuroStoxx50 indices. We consider the Bootstrap particle filter and the Auxiliary particle filter, which we employed in section XX; in addition to these, we add the Liu and West filter, which did not feature in that section of this work.\
In the case of the SP500 index, although the number of particles increases, both RMSE and MAE increase (although by very small amounts) for the Auxiliary Particle Filter, while they slightly decrease for the Bootstrap PF. Instead, for the Liu and West filter we see a sizable deterioration in the performance: this suggests that the parameter learning process embedded in the LWF is actually pointing in a different direction than that of the proxy, suggesting that either the proxy does not capture well the latent volatility, or a SV model may not be the best tool to disclose such latent process.\footnote{Note that at N=1000, the LWF yields a RMSE and a MAE. In words, worse approximations of the target distribution of the LWF are actually best at replicating the proxy for the SP500. A possible interpretation goes, once again, as that reported in the main text.}\
In the case of the EuroStoxx50 index, while the Bootstrap filter slightly increases both RMSE and MAE when N is raised to 50000, the APF shows mild improvements in its MAE performance, while the RMSE criterion points in the opposite direction. The criteria disagree also in the case of the LW filter, as the MAE mildly prefers the performance of N=10000, while the RMSE detects minor improvements for N=50000.
